Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-28 doc/papers/iop2014/paper.tex
Pierrick's review

    In addition to these compilation steps, Pythran provides a C++ runtime 
    library that
    leverages the C++ STL to provide generic containers, and the Numeric 
    % FIXME (phrase trop longuuuuuuuuue) 
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO

    % FIXME  (explain that annotations can be added to the code?)
    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, % FIXME (cut sentence) 
    without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code.

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is then translated to polymorphic C++ code through the
generation of templated classes in place of Python functions. These
classes are either used from regular C++ % FIXME (not clear what do you mean?)
code or instantiated for given types to generate a native Python module.

@@

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.7.2 and clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
In other words, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, i.e. whose return value
only depends on the value of their arguments. This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function 
whether or not this argument may be updated in the function body. These three
analyses work inter-procedurally, as illustrated in the following example:

@@

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because \texttt{print b} involves an \textbf{I/O}.

@@

\texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
is constant, i.e. its value only depends on literals. For instance the
expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
argument is a literal. 
% FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)

\texttt{UseDefChains} is a classical analysis from the static compilation world. For
each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
The result can be used to drive various code transformations, for instance to
remove dead code, such as a \textbf{def} followed by aeither nothing or a 
\textbf{def} is useless. 
It is used in Pythran to avoid false polymorphism. An intuitive way to represent
use-def chains is illustrated on the next code snippet:

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect) % FIXME example?
The fact that the last \textbf{use} is always preceded by
a \textbf{def} that only points to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).
% FIXME The scalar renaming part is unclear

@@

However, there are some information available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. This approach
is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]+i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i]+i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

\begin{lstlisting}
def valid_conversion(n):
    # this map can be converted to imap
    L = map(math.cos, range(n))
    return sum(l) # sum iterates once on its input

@@

Using the proper library, the C++ language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the C++ Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural choice to represent Python's \texttt{long} in C++. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the use of vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregate all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        and hence increase the impact of the two following optimizations.

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...

@@

Although the AVX instruction set makes it possible to store 4 double precision
floats, one does not get a 4x speed up because of the (unaligned) memory transfers
to and from vector registers.

@@

The 4 hyper-threaded cores give an extra performance boost. Unfortunately, the
load is not sufficient to get more than an average 4x speed up compared to the
vectorized version. In the end, Pythran generates a native module that performs
roughly 11 times faster than the original version.

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developers to
be able to take advantage of them.

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    "#omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}

The loop is flagged as parallel, performing a reduction using the \texttt{+}
operator on the \texttt{hits} variable. Variables \texttt{x}, \texttt{y} and
\texttt{dist} are automatically made local by Pythran after a
\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, i.e. local to
a thread and not shared with other threads, by OpenMP.

@@

Next section performs an in-depth comparison of Pythran with for Python
optimizers: Numba, PyPy, ShedSkin and numexpr.

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

@@

The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
for a numerical application. There are two
reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much to gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
hence the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

@@

\begin{lstlisting}
def arc_distance(theta_1, phi_1, theta_2, phi_2):
    """
    Calculates the pairwise arc distance
    between all points in vector a and b.
    """
    temp = (np.sin((theta_2-theta_1)/2)**2
        + np.cos(theta_1)*np.cos(theta_2)
          * np.sin((phi_2-phi_1)/2)**2)
    distance_matrix = 2 * np.arctan2(
            sqrt(temp),sqrt(1-temp))
    return distance_matrix
\end{lstlisting}

@@

It shows a small benefit from using expression templates on their own, most
certainly because the loop control overhead is negligible compared to the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
an overall speedup greater than x11 compared to the original Numpy version, 
without
changing the input code. Almost on the opposite, Numexpr requires rewriting the 
input and does not achieve the same level of performance as Pythran when OpenMP
and AVX are combined.

@@

% FIXME (just thinking about it, did you say what CPyhton?)

>>>>

    In addition to these compilation steps, Pythran provides a C++ runtime 
    library that
    leverages the C++ STL to provide generic containers, and the Numeric 
    % FIXME (phrase trop longuuuuuuuuue) 
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO
    % It is more a classical "use" of the language

    % FIXME  (explain that annotations can be added to the code?)
    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, % FIXME (cut sentence) 
    without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code. 
    % It may be a little too ... untactful?

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is then translated to polymorphic C++ code through the
generation of templated classes in place of Python functions. These
classes are either used to generate regular C++ % FIXME (not clear what do you mean?) PB : Better?
code or instantiated for given types to generate a native Python module.

@@

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.8.2 and clang++ 3.3.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, \ie{they do not assume any type for the variables manipulated
by the program}. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
In other words, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

@@

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, \ie{whose return value
only depends from the value of their arguments and do not change any global state}.
This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function 
whether or not this argument may be updated in the function body. These three
analyses work inter-procedurally, as illustrated in the following example:

@@

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because \texttt{print b} involves an \textbf{I/O}.
% Which is not really usefull as there is no way to say pythran we want to execut it in parallel
% (but logging information is still displayed)

@@

\texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
is constant, \ie{its value only depends on literals}. For instance the
expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
argument is a literal. 
% FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)
% PB : explain later I think.

\texttt{UseDefChains} is a classical analysis from the static compilation world. For
each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
The result can be used to drive various code transformations, for instance to
remove dead code, such as a \textbf{def} followed by either nothing or a 
\textbf{def} is useless. 
It is used in Pythran to avoid false polymorphism. An intuitive way to represent
use-def chains is illustrated on the next code snippet:

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect). % FIXME example?
In the use-def Chains, we can gather all nodes tied with a \textbf{use} node or
preceded by a common \textbf{def} node.
All these nodes refere to the same variable and can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming). It avoid false polymorphism.
% FIXME The scalar renaming part is unclear
% PB : Is it clearer now? I don't speak about UD node but we don't care, do we?

@@

However, there are some information available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. This approach
is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]i + i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i] + i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

\begin{lstlisting}
def valid_conversion(n):
    # this map can be converted to imap
    L = map(math.cos, range(n))
    return sum(L) # sum iterates once on its input

@@

Using the proper library, the \cpp language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the \cpp Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the use of vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregate all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        and hence increase the impact of the two following optimizations.
        % what about binary func?

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
% PB : Not sure it is needed as audiance don't really care about it.

@@

Although the AVX instruction set makes it possible to store 4 double precision
floats, one does not get a 4x speed up because of the (unaligned) memory transfers
to and from vector registers.
% 1.33x speedup is not really good if the best expected is 4x. unaligned explain why it is not better 
% than 2x but what about the others 0.66?

@@

The 4 hyper-threaded cores give an extra performance boost. Unfortunately, the
load is not sufficient to get more than an average 4x speed up compared to the
vectorized version.
% You complain about 4x because you only have 3.98x ???
In the end, Pythran generates a native module that performs
roughly 11\times faster than the original version.

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developers to
be able to take advantage of them.
%Does python programmer really use it? Or people using it may write their native code?
% They may write it using python only to prototyping?

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    "#omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}
% why #? I thought string was enough?

The loop is flagged as parallel, performing a reduction using the \texttt{+}
operator on the \texttt{hits} variable. Variables \texttt{x}, \texttt{y} and
\texttt{dist} are automatically made local by Pythran after a
\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, \ie{local to
a thread and not shared with other threads, by OpenMP}.

@@

Next section performs an in-depth comparison of Pythran with others Python
optimizers: Numba, PyPy, ShedSkin and numexpr.

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.8.2, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

@@

The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
for a numerical application. There are two
reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much to gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
hence the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.
% Showing parallelization AND np.array at the same time confuse reader as we
% don't know how much each optimisation contribute in the speed up

@@

\begin{lstlisting}
def arc_distance(theta_1, phi_1, theta_2, phi_2):
    """
    Calculates the pairwise arc distance
    between all points in vector a and b.
    """
    temp = (np.sin((theta_2 - theta_1) / 2) ** 2
        + np.cos(theta_1) * np.cos(theta_2)
          * np.sin((phi_2 - phi_1) / 2) ** 2)
    distance_matrix = 2 * np.arctan2(
            sqrt(temp), sqrt(1 - temp))
    return distance_matrix
\end{lstlisting}

@@

It shows a small benefit from using expression templates on their own, most
% In fact there are small benefits for compiling this code using only pythran without AVX and openMP.
certainly because the loop control overhead is negligible compared to the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
an overall speedup greater than x11 compared to the original Numpy version, 
without
changing the input code. Almost on the opposite, Numexpr requires rewriting the 
input and does not achieve the same level of performance as Pythran when OpenMP
and AVX are combined.

@@

% FIXME (just thinking about it, did you say what CPyhton?)
% PB : I think it is ok as audience is Python programmers
